---
title: "Multiome Analysis: JUND KD Only (NTC + JUND_KD)"
author: "Yonatan Degefu"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 6)
```

# Overview

Re-analysis of the 10x Multiome dataset using only **3 samples** (NTC_rep1,
NTC_rep2, JUND_KD), so that JUNB_KD does not influence WNN clustering, Tsoi
labeling, or any downstream analysis. Reuses per-sample checkpoints from the
full 4-sample pipeline to skip redundant computation.


# 1. Setup

```{r libraries}
suppressPackageStartupMessages({
  library(Seurat)
  library(Signac)
  library(UCell)
  library(EnsDb.Hsapiens.v86)
  library(GenomicRanges)
  library(GenomeInfoDb)
  library(Rsamtools)
  library(SingleCellExperiment)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(patchwork)
  library(tibble)
})

set.seed(42)
options(future.globals.maxSize = 8000 * 1024^2)

run_opts <- list(
  tss_fast = TRUE,
  use_checkpoints = TRUE,
  force_rerun = FALSE
)
```

```{r paths}
base_path <- "/Volumes/FallahiLab/Maize-Data/Data/Magda_Bujnowska/20240530_multiome_analysis/20240720_Multiome_analysis"
tsoi_path <- "/Volumes/FallahiLab/Maize-Data/Data/Yonatan_Degefu/ap1_multiomic_analysis/data/processed/curated_gene_sets/Tsoi_et_al_gene_list.csv"

# if you want to rerun and reload the merged file uncomment this
# merged <- readRDS(
#   "analysis/sc_rna_atac_analysis/results/jund_only/merged_multiome_jund_only.rds"
# )


# Output goes to a separate subdirectory so nothing overwrites the 4-sample results
output_dir <- "/Volumes/FallahiLab/Maize-Data/Data/Yonatan_Degefu/ap1_multiomic_analysis/analysis/sc_rna_atac_analysis/results/jund_only"
fig_dir <- file.path(output_dir, "figures")
qc_dir <- file.path(output_dir, "qc")
tbl_dir <- file.path(output_dir, "tables")
ckpt_dir <- file.path(output_dir, "checkpoints")

for (d in c(output_dir, fig_dir, qc_dir, tbl_dir, ckpt_dir)) {
  dir.create(d, recursive = TRUE, showWarnings = FALSE)
}

# Path to the EXISTING 4-sample checkpoints (we reuse these)
shared_ckpt <- "/Volumes/FallahiLab/Maize-Data/Data/Yonatan_Degefu/ap1_multiomic_analysis/analysis/sc_rna_atac_analysis/results/checkpoints"

# 3-sample subset
sample_names <- c("NTC_rep1", "NTC_rep2", "JUND_KD")
sample_dirs <- c(
  NTC_rep1 = file.path(base_path, "sample1_count_output/outs"),
  NTC_rep2 = file.path(base_path, "sample2_count_output/outs"),
  JUND_KD = file.path(base_path, "sample4_count_output/outs")
)

condition_colors <- c(
  NTC_rep1 = "#5A5A5A",
  NTC_rep2 = "#AAAAAA",
  JUND_KD = "#D62728"
)


tsoi_colors_7 <- c(
  Melanocytic = "#2E5A88",
  Transitory_Melanocytic = "#5B8DBF",
  Transitory = "#F4A261",
  NC_Transitory = "#A8D08D",
  NeuralCrest = "#4F8C5A",
  Undiff_NeuralCrest = "#D87575",
  Undifferentiated = "#A8324A"
)

tsoi_colors_4 <- c(
  Melanocytic = "#4E79A7",
  Transitory = "#F28E2B",
  NeuralCrest = "#59A14F",
  Undifferentiated = "#E15759",
  `Melanocytic (low conf)` = "#A8C2D8",
  `Transitory (low conf)` = "#F8C994",
  `NeuralCrest (low conf)` = "#AECEA9",
  `Undifferentiated (low conf)` = "#F0AAAB"
)
```

```{r qc-thresholds}
qc_thresholds <- list(
  # RNA
  nFeature_RNA_min       = 200,
  nFeature_RNA_max       = 8000,
  nCount_RNA_max         = 30000,
  percent_mt_max         = 15,
  # ATAC
  nCount_ATAC_min        = 3000,
  nCount_ATAC_max        = 100000,
  pct_reads_in_peaks_min = 15,
  blacklist_ratio_max    = 0.05,
  TSS_enrichment_min     = 2,
  nucleosome_signal_max  = 4,
  # Doublets
  doublet_remove         = TRUE
)
```

# 2. Load from existing checkpoints

Per-sample objects (RNA + doublet calls) and ATAC quantification were computed
in the full 4-sample pipeline. We load those and drop JUNB_KD.

```{r load-checkpoints}
per_sample <- readRDS(file.path(shared_ckpt, "02_per_sample_with_doublets.rds"))
per_sample[["JUNB_KD"]] <- NULL
cat("Loaded per-sample objects for:", paste(names(per_sample), collapse = ", "), "\n")
sapply(per_sample, ncol)

atac_quant <- readRDS(file.path(shared_ckpt, "03_atac_quant.rds"))
atac_quant$atac_counts_list[["JUNB_KD"]] <- NULL
# Fragment objects are an unnamed list ordered: NTC_rep1, NTC_rep2, JUNB_KD, JUND_KD
# Keep indices 1, 2, 4 (drop JUNB_KD at index 3)
atac_quant$frag_objs <- atac_quant$frag_objs[c(1, 2, 4)]
cat("ATAC counts samples:", paste(names(atac_quant$atac_counts_list), collapse = ", "), "\n")
```

# 3. Doublet summary

```{r doublet-summary}
doublet_summary <- data.frame(
  sample   = sample_names,
  n_cells  = sapply(per_sample, ncol),
  n_doublet = sapply(per_sample, function(o) sum(o$scDblFinder_class == "doublet")),
  pct_doublet = sapply(per_sample, function(o)
    round(100 * mean(o$scDblFinder_class == "doublet"), 2))
)
print(doublet_summary)
write.csv(doublet_summary, file.path(qc_dir, "doublet_summary.csv"), row.names = FALSE)
```

# 4. Diagnostic QC plots (no filtering yet)

```{r qc-build-long-df}
qc_long <- bind_rows(lapply(per_sample, function(o) {
  o@meta.data %>%
    tibble::rownames_to_column("cell") %>%
    dplyr::select(cell, orig.ident,
           nFeature_RNA, nCount_RNA, percent.mt, percent.ribo,
           atac_fragments, atac_peak_region_fragments,
           pct_reads_in_peaks_arc, scDblFinder_score, scDblFinder_class)
})) %>%
  mutate(orig.ident = factor(orig.ident, levels = sample_names))
```

```{r qc-violin-plots, fig.width=12, fig.height=10}
make_violin <- function(df, ycol, ylab, hlines = NULL, log_y = FALSE) {
  p <- ggplot(df, aes(orig.ident, .data[[ycol]], fill = orig.ident)) +
    geom_violin(scale = "width", alpha = 0.7) +
    geom_jitter(width = 0.15, size = 0.15, alpha = 0.2) +
    scale_fill_manual(values = condition_colors) +
    labs(x = NULL, y = ylab) +
    theme_classic(base_size = 12) +
    theme(legend.position = "none")
  if (log_y) p <- p + scale_y_log10()
  if (!is.null(hlines)) {
    p <- p + geom_hline(yintercept = hlines, linetype = "dashed",
                        color = "red", alpha = 0.7)
  }
  p
}

rna_qc_plots <- list(
  make_violin(qc_long, "nFeature_RNA", "nFeature_RNA",
              c(qc_thresholds$nFeature_RNA_min, qc_thresholds$nFeature_RNA_max)),
  make_violin(qc_long, "nCount_RNA", "nCount_RNA (log)",
              c(qc_thresholds$nCount_RNA_max), log_y = TRUE),
  make_violin(qc_long, "percent.mt", "% mitochondrial",
              c(qc_thresholds$percent_mt_max)),
  make_violin(qc_long, "percent.ribo", "% ribosomal"),
  make_violin(qc_long, "scDblFinder_score", "doublet score")
)

rna_panel <- wrap_plots(rna_qc_plots, ncol = 3)
print(rna_panel)
ggsave(file.path(qc_dir, "QC_pre_filter_violins.pdf"),
       rna_panel, width = 12, height = 8)
```

```{r qc-scatter-tss-vs-frags, fig.width=10, fig.height=4}
p_scatter <- ggplot(qc_long, aes(nCount_RNA, scDblFinder_score, color = scDblFinder_class)) +
  geom_point(size = 0.3, alpha = 0.3) +
  facet_wrap(~orig.ident, ncol = 3) +
  scale_x_log10() +
  labs(title = "Doublet score vs library size") +
  theme_classic(base_size = 11)
print(p_scatter)
ggsave(file.path(qc_dir, "QC_doublet_scatter.pdf"), p_scatter, width = 10, height = 4)
```

# 5. Apply RNA QC + doublet filter

```{r apply-rna-filter}
valid_barcodes <- list()
for (s in sample_names) {
  obj <- per_sample[[s]]
  keep <- obj$nFeature_RNA >= qc_thresholds$nFeature_RNA_min &
          obj$nFeature_RNA <= qc_thresholds$nFeature_RNA_max &
          obj$nCount_RNA   <= qc_thresholds$nCount_RNA_max &
          obj$percent.mt   <= qc_thresholds$percent_mt_max
  if (qc_thresholds$doublet_remove) {
    keep <- keep & obj$scDblFinder_class != "doublet"
  }
  valid_barcodes[[s]] <- colnames(obj)[keep]
  per_sample[[s]] <- subset(obj, cells = valid_barcodes[[s]])
  cat(s, ":", length(valid_barcodes[[s]]), "/", ncol(obj), "cells kept\n")
}

pre_filter_counts <- doublet_summary$n_cells  # from section 3, before any filtering
rna_qc_summary <- data.frame(
  sample      = sample_names,
  pre_filter  = pre_filter_counts,
  post_filter = sapply(per_sample, ncol)
)
rna_qc_summary$pct_kept <- round(100 * rna_qc_summary$post_filter /
                                   rna_qc_summary$pre_filter, 1)
print(rna_qc_summary)
write.csv(rna_qc_summary, file.path(qc_dir, "rna_qc_summary.csv"), row.names = FALSE)
```

## 5b. Post-filter QC violins

```{r qc-post-filter-violins, fig.width=12, fig.height=8}
qc_long_post <- bind_rows(lapply(sample_names, function(s) {
  per_sample[[s]]@meta.data %>%
    tibble::rownames_to_column("cell") %>%
    dplyr::select(cell, orig.ident, nFeature_RNA, nCount_RNA,
                  percent.mt, percent.ribo)
})) %>%
  mutate(orig.ident = factor(orig.ident, levels = sample_names))

post_plots <- list(
  make_violin(qc_long_post, "nFeature_RNA", "nFeature_RNA (post-filter)"),
  make_violin(qc_long_post, "nCount_RNA", "nCount_RNA (post-filter, log)", log_y = TRUE),
  make_violin(qc_long_post, "percent.mt", "% mito (post-filter)"),
  make_violin(qc_long_post, "percent.ribo", "% ribo (post-filter)")
)
post_panel <- wrap_plots(post_plots, ncol = 2)
print(post_panel)
ggsave(file.path(qc_dir, "QC_post_filter_violins.pdf"),
       post_panel, width = 10, height = 7)

# Distribution summary
post_dist <- qc_long_post %>%
  group_by(orig.ident) %>%
  summarise(
    median_nFeat = median(nFeature_RNA),
    median_nCount = median(nCount_RNA),
    median_mt = round(median(percent.mt), 2),
    .groups = "drop"
  )
print(post_dist)
write.csv(post_dist, file.path(qc_dir, "rna_qc_post_filter_distribution.csv"),
          row.names = FALSE)
```

# 8. Merge RNA and build ChromatinAssay

ATAC quantification was loaded from checkpoint 03. The union peak set includes
peaks from all 4 original samples (pragmatic approach — extra peaks are filtered
by `min.cells = 10` in `CreateChromatinAssay`).

```{r extract-atac}
atac_counts_list <- atac_quant$atac_counts_list
frag_objs        <- atac_quant$frag_objs

sapply(atac_counts_list, ncol)

shared_peaks <- Reduce(intersect, lapply(atac_counts_list, rownames))
atac_counts_list <- lapply(atac_counts_list, function(m) m[shared_peaks, , drop = FALSE])
atac_counts <- do.call(cbind, atac_counts_list)
dim(atac_counts)
```

```{r merge-and-build-atac}
annotations <- GetGRangesFromEnsDb(ensdb = EnsDb.Hsapiens.v86)
seqlevelsStyle(annotations) <- "UCSC"
genome(annotations) <- "hg38"

merged <- merge(per_sample[[1]],
                y = per_sample[2:3],
                add.cell.ids = sample_names,
                project = "AP1_multiome_JUND")
merged <- JoinLayers(merged)

common_cells <- intersect(colnames(merged), colnames(atac_counts))
length(common_cells)
merged      <- subset(merged, cells = common_cells)
atac_counts <- atac_counts[, common_cells]

chrom_assay <- CreateChromatinAssay(
  counts     = atac_counts,
  sep        = c(":", "-"),
  genome     = "hg38",
  fragments  = unname(frag_objs),
  annotation = annotations,
  min.cells  = 10
)
merged[["ATAC"]] <- chrom_assay
```

# 8b. RNA gene-level detection diagnostic

```{r gene-detection-distribution, fig.width=10, fig.height=4}
DefaultAssay(merged) <- "RNA"
rna_counts_full <- GetAssayData(merged, assay = "RNA", layer = "counts")

gene_detection <- rowMeans(rna_counts_full > 0)

gene_total <- Matrix::rowSums(rna_counts_full)
gene_n_detected <- Matrix::rowSums(rna_counts_full > 0)
gene_mean_when_detected <- ifelse(
  gene_n_detected > 0,
  gene_total / gene_n_detected,
  0
)
names(gene_mean_when_detected) <- rownames(rna_counts_full)

gene_qc_df <- data.frame(
  gene = rownames(rna_counts_full),
  pct_cells_detected = 100 * gene_detection,
  mean_count_when_detected = gene_mean_when_detected
)
write.csv(gene_qc_df,
          file.path(qc_dir, "rna_gene_detection_distribution.csv"),
          row.names = FALSE)

detection_bins <- data.frame(
  threshold_pct_cells = c(0, 1, 5, 10, 25, 50),
  n_genes_at_or_above = sapply(c(0, 1, 5, 10, 25, 50), function(t) {
    sum(gene_qc_df$pct_cells_detected >= t)
  })
)
detection_bins$pct_of_total <- round(
  100 * detection_bins$n_genes_at_or_above / nrow(gene_qc_df), 1
)
print(detection_bins)
write.csv(detection_bins,
          file.path(qc_dir, "rna_gene_detection_bins.csv"),
          row.names = FALSE)

p_detect <- ggplot(gene_qc_df, aes(pct_cells_detected)) +
  geom_histogram(bins = 80, fill = "#4E79A7", color = "white") +
  geom_vline(xintercept = 10, linetype = "dashed", color = "red") +
  annotate("text", x = 12, y = Inf, label = "min.pct = 10%",
           hjust = 0, vjust = 1.5, color = "red") +
  scale_x_continuous(breaks = seq(0, 100, 10)) +
  labs(
    x = "% of cells detecting the gene",
    y = "# genes",
    title = "Per-gene detection rate (RNA assay, JUND-only subset)",
    subtitle = paste0(
      "Total genes in assay: ", nrow(gene_qc_df),
      " | Genes detected in >=10% of cells: ",
      sum(gene_qc_df$pct_cells_detected >= 10)
    )
  ) +
  theme_classic(base_size = 12)

p_when_detected <- ggplot(
  subset(gene_qc_df, pct_cells_detected >= 10),
  aes(mean_count_when_detected)
) +
  geom_histogram(bins = 80, fill = "#59A14F", color = "white") +
  scale_x_log10() +
  geom_vline(xintercept = 2, linetype = "dashed", color = "red") +
  labs(
    x = "Mean UMI count in cells that detect the gene (log10)",
    y = "# genes",
    title = "Expression level when detected (testable universe)",
    subtitle = "Restricted to genes detected in >=10% of cells. Bar at count=2."
  ) +
  theme_classic(base_size = 12)

print(p_detect + p_when_detected)
ggsave(file.path(qc_dir, "QC_gene_detection.pdf"),
       p_detect + p_when_detected, width = 12, height = 4)
```

# 9. ATAC QC

```{r atac-qc-compute}
DefaultAssay(merged) <- "ATAC"
merged <- NucleosomeSignal(merged)
merged <- TSSEnrichment(merged, fast = run_opts$tss_fast)
merged$blacklist_ratio <- FractionCountsInRegion(
  object  = merged,
  assay   = "ATAC",
  regions = blacklist_hg38_unified
)
merged$pct_reads_in_peaks <- 100 * merged$nCount_ATAC /
  (merged$atac_fragments + 1e-9)
```

```{r atac-qc-diagnostic-plots, fig.width=14, fig.height=9}
atac_qc_df <- merged@meta.data %>%
  tibble::rownames_to_column("cell") %>%
  mutate(orig.ident = factor(orig.ident, levels = sample_names))

atac_qc_plots <- list(
  make_violin(atac_qc_df, "nCount_ATAC", "nCount_ATAC (log)",
              c(qc_thresholds$nCount_ATAC_min, qc_thresholds$nCount_ATAC_max),
              log_y = TRUE),
  make_violin(atac_qc_df, "TSS.enrichment", "TSS enrichment",
              c(qc_thresholds$TSS_enrichment_min)),
  make_violin(atac_qc_df, "nucleosome_signal", "nucleosome signal",
              c(qc_thresholds$nucleosome_signal_max)),
  make_violin(atac_qc_df, "blacklist_ratio", "blacklist ratio",
              c(qc_thresholds$blacklist_ratio_max)),
  make_violin(atac_qc_df, "pct_reads_in_peaks", "% reads in peaks (Signac)",
              c(qc_thresholds$pct_reads_in_peaks_min))
)

atac_qc_panel <- wrap_plots(atac_qc_plots, ncol = 3)
print(atac_qc_panel)
ggsave(file.path(qc_dir, "QC_atac_pre_filter_violins.pdf"),
       atac_qc_panel, width = 14, height = 9)
```

```{r atac-fragment-histogram, fig.width=12, fig.height=4}
p_frag <- FragmentHistogram(merged, group.by = "orig.ident")
print(p_frag)
ggsave(file.path(qc_dir, "QC_fragment_histogram.pdf"),
       p_frag, width = 12, height = 4)
```

```{r atac-density-scatter, fig.width=10, fig.height=8}
p_dens <- DensityScatter(merged, x = "nCount_ATAC", y = "TSS.enrichment",
                         log_x = TRUE, quantiles = TRUE)
print(p_dens)
ggsave(file.path(qc_dir, "QC_density_TSS_vs_nCount_ATAC.pdf"),
       p_dens, width = 8, height = 6)
```

# 10. Apply ATAC QC

```{r atac-qc-filter}
n_before <- ncol(merged)
merged <- subset(merged,
  subset = nCount_ATAC       > qc_thresholds$nCount_ATAC_min  &
           nCount_ATAC       < qc_thresholds$nCount_ATAC_max  &
           TSS.enrichment    > qc_thresholds$TSS_enrichment_min &
           nucleosome_signal < qc_thresholds$nucleosome_signal_max &
           blacklist_ratio   < qc_thresholds$blacklist_ratio_max  &
           pct_reads_in_peaks > qc_thresholds$pct_reads_in_peaks_min
)
n_after <- ncol(merged)
cat("ATAC QC: kept", n_after, "/", n_before, "cells\n")

cell_counts_summary <- data.frame(
  sample        = sample_names,
  rna_post_qc   = sapply(sample_names, function(s) length(valid_barcodes[[s]])),
  atac_quant    = sapply(sample_names, function(s) ncol(atac_counts_list[[s]])),
  joint_post_qc = as.integer(table(factor(merged$orig.ident, levels = sample_names)))
)
cell_counts_summary$dropped <- cell_counts_summary$rna_post_qc - cell_counts_summary$joint_post_qc
print(cell_counts_summary)
write.csv(cell_counts_summary, file.path(qc_dir, "cell_counts_per_sample.csv"),
          row.names = FALSE)
```

# 11. Normalize, embed (RNA + ATAC)

```{r normalize-rna}
DefaultAssay(merged) <- "RNA"
merged <- NormalizeData(merged, verbose = FALSE)
merged <- CellCycleScoring(merged,
                           s.features   = cc.genes.updated.2019$s.genes,
                           g2m.features = cc.genes.updated.2019$g2m.genes,
                           set.ident = FALSE)
merged$CC.Difference <- merged$S.Score - merged$G2M.Score

merged <- SCTransform(merged,
                      vars.to.regress = c("percent.mt", "CC.Difference"),
                      verbose = FALSE)
```

```{r rna-pca-cluster}
merged <- RunPCA(merged, npcs = 30, verbose = FALSE, seed.use = 42)

dims_use <- 1:16
merged <- FindNeighbors(merged, dims = dims_use, verbose = FALSE)

leiden_ok <- tryCatch({
  merged <- FindClusters(merged, algorithm = 4, resolution = 0.5,
                         random.seed = 42, verbose = FALSE)
  TRUE
}, error = function(e) {
  message("Leiden failed: ", conditionMessage(e),
          " — falling back to Louvain (algorithm = 2).")
  FALSE
})
if (!leiden_ok) {
  merged <- FindClusters(merged, algorithm = 2, resolution = 0.5,
                         random.seed = 42, verbose = FALSE)
  warning("CLUSTERING FALLBACK: Leiden was unavailable; used Louvain instead.",
          immediate. = TRUE)
}

merged$final_clusters <- merged$seurat_clusters
merged <- RunUMAP(merged, dims = dims_use, seed.use = 42,
                  reduction.name = "umap.rna", verbose = FALSE)
```

```{r atac-embed}
DefaultAssay(merged) <- "ATAC"
merged <- RunTFIDF(merged)
merged <- FindTopFeatures(merged, min.cutoff = "q5")
merged <- RunSVD(merged, reduction.name = "lsi")
merged <- RunUMAP(merged, reduction = "lsi", dims = 2:30,
                  reduction.name = "umap.atac", seed.use = 42, verbose = FALSE)
DefaultAssay(merged) <- "SCT"
```

## Elbow plots for dimensionality selection

Inspect these before proceeding to WNN. Adjust `pca_dims` and `lsi_dims`
in the next chunk based on where the curves flatten.

```{r elbow-plots, fig.width=12, fig.height=5}
# PCA elbow
p_pca_elbow <- ElbowPlot(merged, ndims = 30, reduction = "pca") +
  ggtitle("PCA elbow plot (RNA)") +
  geom_vline(xintercept = 16, linetype = "dashed", color = "red", alpha = 0.6) +
  annotate("text", x = 17, y = Inf, label = "default: 16",
           hjust = 0, vjust = 1.5, color = "red", size = 3)

# LSI variance plot (drop dim 1 — correlates with depth)
lsi_stdev <- Stdev(merged[["lsi"]])
lsi_df <- data.frame(
  component = seq_along(lsi_stdev),
  stdev = lsi_stdev
)
p_lsi_elbow <- ggplot(lsi_df[lsi_df$component >= 2, ],
                      aes(component, stdev)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 30, linetype = "dashed", color = "red", alpha = 0.6) +
  annotate("text", x = 31, y = Inf, label = "default: 30",
           hjust = 0, vjust = 1.5, color = "red", size = 3) +
  labs(x = "LSI component", y = "Standard deviation",
       title = "LSI elbow plot (ATAC, dim 1 excluded)") +
  theme_classic(base_size = 12)

print(p_pca_elbow + p_lsi_elbow)
ggsave(file.path(fig_dir, "PCA_LSI_elbow_plots.pdf"),
       p_pca_elbow + p_lsi_elbow, width = 12, height = 5)
```

# 12. WNN: joint RNA + ATAC neighbor graph, UMAP, clustering

```{r wnn-build}
# Edit these after inspecting elbow plots above
pca_dims <- 1:16
lsi_dims <- 2:25

merged <- FindMultiModalNeighbors(
  merged,
  reduction.list = list("pca", "lsi"),
  dims.list      = list(pca_dims, lsi_dims),
  modality.weight.name = "RNA.weight",
  verbose = FALSE
)

merged <- RunUMAP(
  merged, nn.name = "weighted.nn",
  reduction.name = "umap.wnn",
  reduction.key  = "wnnUMAP_",
  seed.use = 42, verbose = FALSE
)
```

## Resolution sweep for leiden

```{r wnn-resolution-sweep, fig.width=14, fig.height=10}
res_grid <- c(0.2, 0.3, 0.4, 0.5,0.6, 0.7, 0.9)

for (r in res_grid) {
  merged <- FindClusters(
    merged, graph.name = "wsnn",
    algorithm = 4,
    resolution = r,
    random.seed = 42, verbose = FALSE
  )
  merged[[paste0("wnn_res_", r)]] <- merged$seurat_clusters
}

sweep_summary <- data.frame(
  resolution = res_grid,
  n_clusters = sapply(res_grid, function(r) {
    length(unique(merged[[paste0("wnn_res_", r)]][, 1]))
  })
)
print(sweep_summary)
write.csv(sweep_summary, file.path(tbl_dir, "wnn_resolution_sweep.csv"),
          row.names = FALSE)

sweep_plots <- lapply(res_grid, function(r) {
  DimPlot(merged, reduction = "umap.wnn",
          group.by = paste0("wnn_res_", r),
          label = TRUE, repel = TRUE) +
    ggtitle(paste0("res = ", r, "  (",
                   length(unique(merged[[paste0("wnn_res_", r)]][, 1])),
                   " clusters)")) +
    NoLegend()
})
sweep_panel <- wrap_plots(sweep_plots, ncol = 3)
print(sweep_panel)
ggsave(file.path(fig_dir, "WNN_resolution_sweep.pdf"),
       sweep_panel, width = 16, height = 10)
```

## Pick a resolution

Inspect `WNN_resolution_sweep.pdf` and set `chosen_res` below.

```{r wnn-final-clusters}
chosen_res <- 0.3   # <-- edit after viewing sweep

merged$wnn_clusters <- merged[[paste0("wnn_res_", chosen_res)]][, 1]
Idents(merged) <- "wnn_clusters"
cat("Using resolution", chosen_res, "with",
    length(unique(merged$wnn_clusters)), "clusters\n")
```

# 13. Tsoi signature scoring (annotation)

```{r ucell}
tsoi_raw <- read.csv(tsoi_path, check.names = FALSE, stringsAsFactors = FALSE)

clean_genes <- function(x) {
  x <- x[!is.na(x) & nchar(trimws(x)) > 0]
  unique(trimws(x))
}

tsoi_combined <- list(
  Melanocytic = clean_genes(tsoi_raw[["Melanocytic"]]),
  Transitory_Melanocytic = clean_genes(tsoi_raw[["Transitory-Melanocytic"]]),
  Transitory = clean_genes(tsoi_raw[["Transitory"]]),
  NC_Transitory = clean_genes(tsoi_raw[["Neural crest-like-Transitory"]]),
  NeuralCrest = clean_genes(tsoi_raw[["Neural crest-like"]]),
  Undiff_NeuralCrest = clean_genes(tsoi_raw[[
    "Undifferentiated-Neural crest-like"
  ]]),
  Undifferentiated = clean_genes(tsoi_raw[["Undifferentiated"]])
)
sapply(tsoi_combined, length)

rna_genes <- rownames(merged[["RNA"]])
tsoi_overlap <- data.frame(
  signature = names(tsoi_combined),
  n_genes = sapply(tsoi_combined, length),
  n_in_rna = sapply(tsoi_combined, function(g) sum(g %in% rna_genes)),
  pct_in_rna = sapply(tsoi_combined, function(g) {
    round(100 * mean(g %in% rna_genes), 1)
  })
)
print(tsoi_overlap)
write.csv(tsoi_overlap, file.path(qc_dir, "tsoi_gene_overlap.csv"),
          row.names = FALSE)

# Filter signatures to genes detected in >=5% of cells
DefaultAssay(merged) <- "RNA"
rna_counts <- GetAssayData(merged, assay = "RNA", layer = "counts")
detection_threshold <- 0.05

tsoi_combined_filt <- lapply(tsoi_combined, function(g) {
  g_in_rna <- intersect(g, rownames(rna_counts))
  if (length(g_in_rna) == 0) return(character(0))
  det <- rowMeans(rna_counts[g_in_rna, , drop = FALSE] > 0)
  g_in_rna[det >= detection_threshold]
})

tsoi_filter_summary <- data.frame(
  signature = names(tsoi_combined),
  n_original = sapply(tsoi_combined, length),
  n_in_rna = sapply(tsoi_combined, function(g) sum(g %in% rna_genes)),
  n_filtered = sapply(tsoi_combined_filt, length)
)
tsoi_filter_summary$pct_retained_vs_in_rna <- round(
  100 * tsoi_filter_summary$n_filtered / pmax(1, tsoi_filter_summary$n_in_rna), 1
)
print(tsoi_filter_summary)
write.csv(tsoi_filter_summary, file.path(qc_dir, "tsoi_signature_filter_summary.csv"),
          row.names = FALSE)

keep_sigs <- names(tsoi_combined_filt)[sapply(tsoi_combined_filt, length) >= 10]
tsoi_combined_filt <- tsoi_combined_filt[keep_sigs]
cat("Signatures retained for scoring:", paste(keep_sigs, collapse = ", "), "\n")

# Score with FILTERED 7-state signatures
DefaultAssay(merged) <- "RNA"
merged <- AddModuleScore_UCell(
  merged,
  features = tsoi_combined_filt,
  name = "_UCellF",
  assay = "RNA"
)

# 4-state collapsed signatures
collapse_map <- c(
  Melanocytic = "Melanocytic",
  Transitory_Melanocytic = "Melanocytic",
  Transitory = "Transitory",
  NC_Transitory = "Transitory",
  NeuralCrest = "NeuralCrest",
  Undiff_NeuralCrest = "NeuralCrest",
  Undifferentiated = "Undifferentiated"
)

tsoi4_combined_filt <- list(
  Melanocytic = unique(unlist(tsoi_combined_filt[
    names(collapse_map)[collapse_map == "Melanocytic"]
  ])),
  Transitory = unique(unlist(tsoi_combined_filt[
    names(collapse_map)[collapse_map == "Transitory"]
  ])),
  NeuralCrest = unique(unlist(tsoi_combined_filt[
    names(collapse_map)[collapse_map == "NeuralCrest"]
  ])),
  Undifferentiated = unique(unlist(tsoi_combined_filt[
    names(collapse_map)[collapse_map == "Undifferentiated"]
  ]))
)
tsoi4_combined_filt <- tsoi4_combined_filt[
  sapply(tsoi4_combined_filt, length) >= 10
]

cat("4-state signature sizes after filtering:\n")
print(sapply(tsoi4_combined_filt, length))

merged <- AddModuleScore_UCell(
  merged,
  features = tsoi4_combined_filt,
  name = "_UCell4",
  assay = "RNA"
)
DefaultAssay(merged) <- "SCT"

# Per-cluster z-score heatmap, 4-state
ucell4_cols <- paste0(names(tsoi4_combined_filt), "_UCell4")
ucell4_mat <- as.data.frame(merged@meta.data[, ucell4_cols, drop = FALSE])
colnames(ucell4_mat) <- names(tsoi4_combined_filt)
ucell4_z <- scale(as.matrix(ucell4_mat))

cluster_tsoi4_mean <- aggregate(
  ucell4_z,
  by = list(cluster = merged$wnn_clusters),
  FUN = mean
)
print(cluster_tsoi4_mean)
write.csv(cluster_tsoi4_mean,
          file.path(tbl_dir, "wnn_cluster_tsoi4_mean_zscore_filtered.csv"),
          row.names = FALSE)

mat4 <- as.matrix(cluster_tsoi4_mean[, -1])
rownames(mat4) <- paste0("c", cluster_tsoi4_mean$cluster)

heat4_df <- as.data.frame(mat4) %>%
  tibble::rownames_to_column("cluster") %>%
  pivot_longer(-cluster, names_to = "signature", values_to = "z")
heat4_df$signature <- factor(heat4_df$signature, levels = names(tsoi4_combined_filt))
heat4_df$cluster <- factor(
  heat4_df$cluster,
  levels = paste0("c", sort(unique(cluster_tsoi4_mean$cluster)))
)



# 7-state heatmap
ucell7f_cols <- paste0(names(tsoi_combined_filt), "_UCellF")
ucell7f_mat <- as.data.frame(merged@meta.data[, ucell7f_cols, drop = FALSE])
colnames(ucell7f_mat) <- names(tsoi_combined_filt)
ucell7f_z <- scale(as.matrix(ucell7f_mat))

cluster_tsoi7f_mean <- aggregate(
  ucell7f_z,
  by = list(cluster = merged$wnn_clusters),
  FUN = mean
)
write.csv(cluster_tsoi7f_mean,
          file.path(tbl_dir, "wnn_cluster_tsoi7_mean_zscore_filtered.csv"),
          row.names = FALSE)

# Labeling from 7-state
mat7f <- as.matrix(cluster_tsoi7f_mean[, -1])
rownames(mat7f) <- paste0("c", cluster_tsoi7f_mean$cluster)

top_state7 <- colnames(mat7f)[max.col(mat7f, ties.method = "first")]
cluster_label_map7 <- setNames(top_state7, as.character(cluster_tsoi7f_mean$cluster))
label_vec7 <- cluster_label_map7[as.character(merged$wnn_clusters)]
names(label_vec7) <- colnames(merged)
merged <- AddMetaData(merged, metadata = label_vec7, col.name = "wnn_cluster_tsoi7_filt")

# Gap and confidence on 4-state
mat4 <- as.matrix(cluster_tsoi4_mean[, -1])
rownames(mat4) <- paste0("c", cluster_tsoi4_mean$cluster)

top_z_4 <- apply(mat4, 1, max)
second_z_4 <- apply(mat4, 1, function(x) sort(x, decreasing = TRUE)[2])
top_lab_4 <- colnames(mat4)[max.col(mat4, ties.method = "first")]
gap_4 <- top_z_4 - second_z_4

cluster_labels_4 <- data.frame(
  cluster = rownames(mat4),
  top_label = top_lab_4,
  top_z = round(top_z_4, 2),
  second_z = round(second_z_4, 2),
  gap = round(gap_4, 2),
  confident = top_z_4 >= 0.5 & gap_4 >= 0.3
)
print(cluster_labels_4)
write.csv(cluster_labels_4, file.path(tbl_dir, "wnn_cluster_tsoi4_labels_with_gap.csv"),
          row.names = FALSE)

# Write 4-state label to Seurat object
stale_cols <- intersect("wnn_cluster_tsoi", colnames(merged@meta.data))
if (length(stale_cols) > 0) {
  merged@meta.data[, stale_cols] <- NULL
  cat("Dropped stale label column(s):", paste(stale_cols, collapse = ", "), "\n")
}

label_map_4 <- setNames(
  cluster_labels_4$top_label,
  sub("^c", "", cluster_labels_4$cluster)
)
label_vec_4 <- label_map_4[as.character(merged$wnn_clusters)]
names(label_vec_4) <- colnames(merged)
merged <- AddMetaData(merged, metadata = label_vec_4, col.name = "wnn_cluster_tsoi")

# Confidence-suffixed column
confident_map <- setNames(
  cluster_labels_4$confident,
  sub("^c", "", cluster_labels_4$cluster)
)
label_vec_4_conf <- label_vec_4
not_conf <- !confident_map[as.character(merged$wnn_clusters)]
label_vec_4_conf[not_conf] <- paste0(label_vec_4_conf[not_conf], " (low conf)")
merged <- AddMetaData(merged, metadata = label_vec_4_conf,
                      col.name = "wnn_cluster_tsoi_with_conf")

cat("Per-cluster 4-state labels written to merged$wnn_cluster_tsoi:\n")
print(table(merged$wnn_clusters, merged$wnn_cluster_tsoi))
```

### Figure 7 plots

```{r}
# plot tsoi
p_tsoi4_heat <- ggplot(heat4_df, aes(signature, cluster, fill = z)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(
    low = "#4575b4",
    mid = "white",
    high = "#d73027",
    midpoint = 0,
    limits = c(-1.5, 1.5),
    breaks = c(-1.5, 0, 1.5),
    oob = scales::squish
  ) +
  labs(
    x = NULL,
    y = "WNN cluster",
    fill = "Mean UCell\nz-score"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, size = 14),
    axis.text.y = element_text(size = 12),
    aspect.ratio = 6/5
  )
print(p_tsoi4_heat)
ggsave(
  file.path(fig_dir, "WNN_cluster_tsoi4_heatmap_filtered_final.pdf"),
  p_tsoi4_heat,
  width = 4,
  height = 4
)
```



# 14. WNN figures

## Figure W1 — Three-panel WNN UMAP

```{r figW1-wnn-umap-trio, fig.width=18, fig.height=6}
p_w_sample <- DimPlot(
  merged,
  reduction = "umap.wnn",
  group.by = "orig.ident",
  cols = condition_colors
) +
  ggtitle("WNN UMAP by sample")

p_w_clust <- DimPlot(
  merged,
  reduction = "umap.wnn",
  group.by = "wnn_clusters",
  label = TRUE,
  repel = TRUE
) +
  ggtitle(paste0("WNN clusters (res = ", chosen_res, ")")) +
  NoLegend()

p_w_tsoi <- DimPlot(
  merged,
  reduction = "umap.wnn",
  group.by = "wnn_cluster_tsoi",
  cols = tsoi_colors_4
) +
  ggtitle("Per-cluster Tsoi label (4-state)")

# Per-cell Tsoi argmax label
ucell4_cols <- paste0(names(tsoi4_combined_filt), "_UCell4")
cell_scores <- merged@meta.data[, ucell4_cols]
colnames(cell_scores) <- names(tsoi4_combined_filt)
cell_scores_z <- scale(cell_scores)
merged$tsoi4_percell <- colnames(cell_scores_z)[max.col(cell_scores_z, ties.method = "first")]

p_w_tsoi_percell <- DimPlot(merged, reduction = "umap.wnn",
                            group.by = "tsoi4_percell", cols = tsoi_colors_4) +
  ggtitle("Per-cell Tsoi label (4-state)")

# W1a: per-cluster Tsoi labels
# figW1a <- p_w_sample + p_w_clust + p_w_tsoi
# ggsave(file.path(fig_dir, "FigW1a_WNN_UMAP_trio_cluster_labels.pdf"),
#        figW1a, width = 14, height = 4)
# print(p_w_sample / p_w_clust / p_w_tsoi)

# W1b: per-cell Tsoi labels
figW1b <- p_w_sample + p_w_clust + p_w_tsoi_percell
ggsave(file.path(fig_dir, "FigW1b_WNN_UMAP_trio_percell_labels.pdf"),
       figW1b, width = 14, height = 4)
print(p_w_sample / p_w_clust / p_w_tsoi_percell)

# --- Continuous UCell scores on UMAP ---
ucell4_cols <- paste0(names(tsoi4_combined_filt), "_UCell4")

p_ucell_list <- lapply(seq_along(ucell4_cols), function(i) {
  FeaturePlot(
    merged,
    features = ucell4_cols[i],
    reduction = "umap.wnn",
    order = TRUE
  ) +
    scale_x_continuous(breaks = c(-5, -2.5, 0, 2.5, 5)) +
    scale_color_viridis_c(
      option = "magma",
      breaks = scales::breaks_pretty(n = 3)
    ) +
    ggtitle(names(tsoi4_combined_filt)[i]) +
    theme(plot.title = element_text(size = 14, face = "bold"))
})

umap_w <- 14 / 3
umap_h <- 4

ucell_ncol <- 2
ucell_nrow <- ceiling(length(p_ucell_list) / ucell_ncol)

fig_ucell <- wrap_plots(p_ucell_list, ncol = ucell_ncol)

print(fig_ucell)

ggsave(
  file.path(fig_dir, "WNN_UMAP_tsoi4_UCell_continuous.pdf"),
  fig_ucell,
  width = umap_w * ucell_ncol,
  height = umap_h * ucell_nrow
)

# --- MITF and SOX10 expression on UMAP ---
DefaultAssay(merged) <- "RNA"

p_mitf <- FeaturePlot(
  merged,
  features = "MITF",
  reduction = "umap.wnn",
  order = TRUE
) +
  scale_x_continuous(breaks = c(-5, -2.5, 0, 2.5, 5)) +
  scale_color_viridis_c(
    option = "magma",
    breaks = scales::breaks_pretty(n = 3)
  ) +
  ggtitle("MITF")

p_sox10 <- FeaturePlot(
  merged,
  features = "SOX10",
  reduction = "umap.wnn",
  order = TRUE
) +
  scale_x_continuous(breaks = c(-5, -2.5, 0, 2.5, 5)) +
  scale_color_viridis_c(
    option = "magma",
    breaks = scales::breaks_pretty(n = 3)
  ) +
  ggtitle("SOX10")

fig_genes <- p_mitf + p_sox10

print(fig_genes)

ggsave(
  file.path(fig_dir, "WNN_UMAP_MITF_SOX10_expression.pdf"),
  fig_genes,
  width = umap_w * 2,
  height = umap_h
)
```


## Figure W3 — Cluster composition per sample

```{r figW3-cluster-composition, fig.width=8, fig.height=5}
sample_levels <- sample_names

cluster_pct <- as.data.frame(merged@meta.data) %>%
  mutate(orig.ident = factor(orig.ident, levels = sample_levels)) %>%
  count(orig.ident, cluster = wnn_clusters) %>%
  group_by(orig.ident) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

p_comp <- ggplot(cluster_pct, aes(orig.ident, pct, fill = cluster)) +
  geom_col(width = 0.75) +
  labs(x = NULL, y = "% of cells", fill = "WNN cluster",
       title = "Cluster composition per sample (JUND-only)") +
  theme_classic(base_size = 13)
print(p_comp)
ggsave(file.path(fig_dir, "FigW3_cluster_composition.pdf"),
       p_comp, width = 8, height = 5)
write.csv(cluster_pct, file.path(tbl_dir, "wnn_cluster_composition.csv"),
          row.names = FALSE)
```

### Propeller analysis for cell proportions

```{r}
result <- run_propeller_analysis(
  seurat_obj = merged,
  rna_cluster_col = "wnn_clusters", # <-- set to your RNA cluster column
  sample_col = "orig.ident",
  kd_sample = "JUND_KD",
  ntc_samples = c("NTC_rep1", "NTC_rep2"),
  sample_colors = sample_colors # the one from your existing script
)
```


## Figure W4 — Cluster shifts: JUND_KD vs NTC

```{r figW4-cluster-shifts}
library(purrr)
library(ggpubr)

library(purrr)
library(ggpubr)
library(dplyr)
library(tidyr)
library(ggplot2)

# -----------------------------
# User inputs
# -----------------------------

cluster_col <- "wnn_clusters"
sample_col <- "orig.ident"

ntc_samples <- c("NTC_rep1", "NTC_rep2")
kd_sample <- "JUND_KD"

sample_order <- c("NTC_rep1", "NTC_rep2", "JUND_KD")

sample_colors <- c(
  "JUND_KD" = "#B51D0E",
  "NTC_rep1" = "grey30",
  "NTC_rep2" = "grey65"
)

dir.create(tbl_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)

# -----------------------------
# 1. Metadata
# -----------------------------

stopifnot(cluster_col %in% colnames(merged@meta.data))
stopifnot(sample_col %in% colnames(merged@meta.data))

meta_df <- merged@meta.data %>%
  as.data.frame() %>%
  mutate(
    cluster = as.character(.data[[cluster_col]]),
    sample = as.character(.data[[sample_col]])
  ) %>%
  dplyr::filter(sample %in% sample_order) %>%
  mutate(
    sample = factor(sample, levels = sample_order),
    cluster = factor(
      cluster,
      levels = sort(unique(as.numeric(as.character(cluster))))
    )
  )

cat("\nCells per sample:\n")
print(table(meta_df$sample))

cat("\nCells per cluster:\n")
print(table(meta_df$cluster))

# -----------------------------
# 2. Percent composition per sample
# -----------------------------

cluster_counts <- meta_df %>%
  dplyr::count(sample, cluster, name = "n_cells") %>%
  group_by(sample) %>%
  mutate(
    sample_total = sum(n_cells),
    pct = 100 * n_cells / sample_total
  ) %>%
  ungroup()

write.csv(
  cluster_counts,
  file.path(tbl_dir, "wnn_cluster_counts_and_pct_by_sample.csv"),
  row.names = FALSE
)

# -----------------------------
# 3. Effect-size table: KD vs each NTC replicate
# -----------------------------

clusters <- levels(meta_df$cluster)

pct_wide <- cluster_counts %>%
  select(sample, cluster, pct) %>%
  pivot_wider(names_from = sample, values_from = pct, values_fill = 0)

effect_sizes <- pct_wide %>%
  mutate(
    mean_NTC_pct = (NTC_rep1 + NTC_rep2) / 2,
    delta_pct = JUND_KD - mean_NTC_pct,
    log2_ratio = log2((JUND_KD + 0.01) / (mean_NTC_pct + 0.01)),
    delta_vs_rep1 = JUND_KD - NTC_rep1,
    delta_vs_rep2 = JUND_KD - NTC_rep2,
    consistent_direction = sign(delta_vs_rep1) == sign(delta_vs_rep2),
    ntc_diff = abs(NTC_rep1 - NTC_rep2),
    exceeds_ntc_noise = abs(delta_pct) > ntc_diff,
    candidate = consistent_direction &
      exceeds_ntc_noise &
      abs(delta_pct) > 1,
    direction = case_when(
      candidate & delta_pct > 0 ~ "enriched in KD",
      candidate & delta_pct < 0 ~ "depleted in KD",
      TRUE ~ ""
    )
  )

# -----------------------------
# 4. Propeller results (load from result$master saved earlier, OR re-run)
# -----------------------------
# Assumes you have 'result' in your environment from run_propeller_analysis()
# If not, source the functions file and run it first.

stopifnot(exists("result"))
stopifnot(!is.null(result$master))

propeller_stats <- result$master %>%
  select(cluster, propeller_p, propeller_fdr, boot_frac_sig_05) %>%
  mutate(cluster = as.character(cluster))

# -----------------------------
# 5. Build annotation tables for the plot
# -----------------------------


format_fdr <- function(fdr) {
  sapply(fdr, function(x) {
    if (is.na(x)) {
      return("")
    }
    # Split into mantissa and exponent
    exp_val <- floor(log10(x))
    mant_val <- x / 10^exp_val
    # Round mantissa to 1 decimal; drop trailing ".0"
    mant_str <- sub("\\.0$", "", sprintf("%.1f", mant_val))
    sprintf("FDR == %s %%*%% 10^%d", mant_str, exp_val)
  })
}

# Format delta as signed percentage (e.g. +11.0%)
format_delta <- function(d) sprintf("%+.1f%%", d)

bracket_df <- propeller_stats %>%
  left_join(
    effect_sizes %>%
      select(cluster, delta_pct, candidate) %>%
      mutate(cluster = as.character(cluster)),
    by = "cluster"
  ) %>%
  left_join(
    cluster_counts %>%
      group_by(cluster) %>%
      summarise(y_max = max(pct), .groups = "drop") %>%
      mutate(cluster = as.character(cluster)),
    by = "cluster"
  ) %>%
  mutate(
    x_center = as.numeric(factor(cluster, levels = clusters)),
    x_ntc = x_center - 0.125, # midpoint of the two NTC bars
    x_kd = x_center + 0.25, # JUND_KD bar
    y_delta = y_max + 1.5, # delta % sits just above tallest bar
    y_bracket = y_max + 4.5, # bracket sits above the delta label
    y_fdr = y_max + 6.5, # FDR sits above the bracket
    delta_label = format_delta(delta_pct),
    fdr_label = format_fdr(propeller_fdr),
    delta_color = ifelse(candidate, "firebrick", "grey40"),
    fdr_color = ifelse(propeller_fdr < 0.05, "firebrick", "grey40")
  )

# Tick marks on each end of the bracket
tick_height <- 0.4

# -----------------------------
# 6. The plot
# -----------------------------

y_top <- max(bracket_df$y_fdr) + 2

p_comp <- ggplot(
  cluster_counts,
  aes(x = cluster, y = pct, fill = sample)
) +
  geom_col(
    position = position_dodge(width = 0.75),
    width = 0.68,
    color = "black",
    linewidth = 0.15
  ) +
  # Delta % label (just above the tallest bar)
  geom_text(
    data = bracket_df,
    aes(x = x_center, y = y_delta, label = delta_label, color = delta_color),
    inherit.aes = FALSE,
    size = 3.8,
    fontface = "bold",
    show.legend = FALSE
  ) +
  # Horizontal bracket: NTC midpoint -> JUND_KD
  geom_segment(
    data = bracket_df,
    aes(x = x_ntc, xend = x_kd, y = y_bracket, yend = y_bracket),
    inherit.aes = FALSE,
    linewidth = 0.4
  ) +
  # Left tick (NTC midpoint)
  geom_segment(
    data = bracket_df,
    aes(x = x_ntc, xend = x_ntc, y = y_bracket - tick_height, yend = y_bracket),
    inherit.aes = FALSE,
    linewidth = 0.4
  ) +
  # Right tick (JUND_KD)
  geom_segment(
    data = bracket_df,
    aes(x = x_kd, xend = x_kd, y = y_bracket - tick_height, yend = y_bracket),
    inherit.aes = FALSE,
    linewidth = 0.4
  ) +
  # FDR label (above the bracket)
  geom_text(
    data = bracket_df,
    aes(x = x_center + 0.06, y = y_fdr, label = fdr_label, color = fdr_color),
    inherit.aes = FALSE,
    size = 3.2,
    fontface = "bold",
    parse = TRUE,
    show.legend = FALSE
  ) +
  scale_color_identity() +
  scale_fill_manual(values = sample_colors) +
  labs(
    x = "WNN cluster",
    y = "% of sample's cells",
    title = "Cell-state composition shifts upon JUND knockdown",
    subtitle = "Percent of cells per WNN cluster. FDR from propeller comparing NTC group vs JUND KD.",
    fill = NULL
  ) +
  coord_cartesian(ylim = c(0, y_top)) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 10, color = "grey30")
  )

print(p_comp)

ggsave(
  file.path(fig_dir, "JUND_WNN_cluster_frequency_with_propeller_fdr.pdf"),
  p_comp,
  width = 11,
  height = 6.5
)

# Also save the propeller stats table alongside the plot for the figure legend
write.csv(
  result$master,
  file.path(tbl_dir, "JUND_WNN_propeller_master.csv"),
  row.names = FALSE
)
```



#  Save joint object and session info

```{r save}
saveRDS(merged, file.path(output_dir, "merged_multiome_jund_only.rds"))

sink(file.path(output_dir, "session_info.txt"))
print(sessionInfo())
sink()
```

# FindAllMarkers: RNA + ATAC, one-vs-all per WNN cluster

```{r find-all-markers}
stopifnot("wnn_clusters" %in% colnames(merged@meta.data))
Idents(merged) <- "wnn_clusters"
cat("Clusters present:\n")
print(table(Idents(merged)))

RNA_STRICT <- 1.0
RNA_LOOSE <- 0.5
ATAC_STRICT <- 0.75
ATAC_LOOSE <- 0.5
PADJ <- 0.05

log_path <- file.path(tbl_dir, "FindAllMarkers_progress.log")
log_msg <- function(msg) {
  ts <- format(Sys.time(), "%Y-%m-%d %H:%M:%S")
  cat(paste0("[", ts, "] ", msg, "\n"), file = log_path, append = TRUE)
  cat(paste0("[", ts, "] ", msg, "\n"))
}
cat("", file = log_path)
log_msg("Starting FindAllMarkers pipeline (JUND-only)")

# RNA: one-vs-all
log_msg("RNA FindAllMarkers starting...")
t0 <- Sys.time()

DefaultAssay(merged) <- "RNA"
merged <- NormalizeData(merged, verbose = FALSE)

rna_all_markers <- FindAllMarkers(
  merged,
  assay = "RNA",
  slot = "data",
  test.use = "wilcox",
  only.pos = TRUE,
  min.pct = 0.1,
  logfc.threshold = 0.1,
  verbose = FALSE
)
rna_all_markers$p_val_adj <- p.adjust(rna_all_markers$p_val, method = "BH")

elapsed_rna <- round(as.numeric(difftime(Sys.time(), t0, units = "mins")), 1)
log_msg(paste("RNA FindAllMarkers done in", elapsed_rna, "minutes;",
              nrow(rna_all_markers), "rows"))

write.csv(rna_all_markers, file.path(tbl_dir, "FindAllMarkers_RNA_full.csv"),
          row.names = FALSE)

rna_summary <- rna_all_markers %>%
  filter(p_val_adj < PADJ, avg_log2FC > RNA_STRICT) %>%
  group_by(cluster) %>%
  summarise(
    n_markers_strict = n(),
    top10 = paste(head(gene[order(-avg_log2FC)], 10), collapse = ", "),
    .groups = "drop"
  )
print(rna_summary)
write.csv(rna_summary, file.path(tbl_dir, "FindAllMarkers_RNA_summary_strict.csv"),
          row.names = FALSE)

rna_summary_loose <- rna_all_markers %>%
  filter(p_val_adj < PADJ, avg_log2FC > RNA_LOOSE) %>%
  group_by(cluster) %>%
  summarise(n_markers_loose = n(), .groups = "drop")
write.csv(rna_summary_loose, file.path(tbl_dir, "FindAllMarkers_RNA_summary_loose.csv"),
          row.names = FALSE)

saveRDS(rna_all_markers, file.path(tbl_dir, "FindAllMarkers_RNA_full.rds"))
log_msg("RNA summary saved; starting ATAC...")
```


# plot heatmap of tsoi genes with FindAllMarkers

```{r}
RNA_MARKER_LOGFC_CUTOFF <- 0.5
RNA_MARKER_PADJ_CUTOFF <- 0.05
rna_markers <- readRDS(file.path(tbl_dir, "FindAllMarkers_RNA_full.rds"))
sig_genes <- rna_markers %>%
  filter(
    p_val_adj < RNA_MARKER_PADJ_CUTOFF,
    avg_log2FC > RNA_MARKER_LOGFC_CUTOFF
  ) %>%
  pull(gene) %>%
  unique()

# intersect with tsoi genes
tsoi_in_markers <- lapply(tsoi4_combined_filt, function(gs) {
  intersect(gs, sig_genes)
})
tsoi_df <- stack(tsoi_in_markers)
colnames(tsoi_df) <- c("gene", "tsoi_state")

# force-include key genes even if they didn't pass the logFC cutoff
# assign each to its biological tsoi state
force_include <- data.frame(
  gene = c("SOX10", "MITF", "AXL"),
  tsoi_state = c("Melanocytic", "Melanocytic", "Undifferentiated"),
  stringsAsFactors = FALSE
)
# only add genes that exist in the RNA assay but aren't already in tsoi_df
force_include <- force_include[!force_include$gene %in% tsoi_df$gene, ]
force_include <- force_include[force_include$gene %in% rownames(merged[["RNA"]]), ]
if (nrow(force_include) > 0) {
  tsoi_df <- rbind(tsoi_df, force_include)
}

# mean scaled expression per cluster
DefaultAssay(merged) <- "RNA"
avg_expr <- AverageExpression(
  merged,
  features = tsoi_df$gene,
  group.by = "wnn_clusters",
  assays = "RNA"
)$RNA
# z-score across clusters (rows = genes)
avg_z <- t(scale(t(as.matrix(avg_expr))))

# --- ComplexHeatmap of Tsoi marker genes ---
library(ComplexHeatmap)
library(circlize)

# order genes by tsoi state
tsoi_df$tsoi_state <- factor(tsoi_df$tsoi_state,
                              levels = c("Melanocytic", "Transitory",
                                         "NeuralCrest", "Undifferentiated"))
tsoi_df <- tsoi_df[order(tsoi_df$tsoi_state), ]
avg_z_ordered <- avg_z[tsoi_df$gene, ]

# clamp z-scores for color consistency
avg_z_ordered[avg_z_ordered > 2] <- 2
avg_z_ordered[avg_z_ordered < -2] <- -2

# row annotation: tsoi state color bar
tsoi_state_colors <- c(
  Melanocytic      = "#4E79A7",
  Transitory       = "#F28E2B",
  NeuralCrest      = "#59A14F",
  Undifferentiated = "#E15759"
)
row_ha <- rowAnnotation(
  State = tsoi_df$tsoi_state,
  col = list(State = tsoi_state_colors),
  show_annotation_name = FALSE,
  show_legend = FALSE
)

# select genes to label: top 10 per tsoi state + must-have genes
N_PER_STATE <- 5
must_label <- c("MITF", "SOX10", "AXL")
top_per_state <- rna_markers %>%
  filter(p_val_adj < RNA_MARKER_PADJ_CUTOFF,
         avg_log2FC > RNA_MARKER_LOGFC_CUTOFF,
         gene %in% tsoi_df$gene) %>%
  inner_join(tsoi_df, by = "gene") %>%
  distinct(gene, .keep_all = TRUE) %>%
  group_by(tsoi_state) %>%
  slice_max(avg_log2FC, n = N_PER_STATE, with_ties = FALSE) %>%
  pull(gene) %>%
  unique()
genes_to_label <- unique(c(must_label[must_label %in% tsoi_df$gene], top_per_state))

# anno_mark for non-overlapping gene labels
label_idx <- which(rownames(avg_z_ordered) %in% genes_to_label)
label_names <- rownames(avg_z_ordered)[label_idx]
right_ha <- rowAnnotation(
  genes = anno_mark(
    at = label_idx,
    labels = label_names,
    labels_gp = gpar(fontsize = 12, fontface = "italic"),
    link_width = unit(8, "mm")
  )
)

# rename columns to C1, C2, ...
colnames(avg_z_ordered) <- paste0("C", seq_len(ncol(avg_z_ordered)))

col_fun <- colorRamp2(c(-2, 0, 2), c("#4575b4", "white", "#d73027"))

ht <- Heatmap(
  avg_z_ordered,
  name = "Mean expression\nz-score",
  col = col_fun,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_names = FALSE,
  column_names_gp = gpar(fontsize = 14),
  width = unit(4.5, "cm"),
  row_split = tsoi_df$tsoi_state,
  row_gap = unit(2, "mm"),
  left_annotation = row_ha,
  right_annotation = right_ha,
  row_title_gp = gpar(fontsize = 12),
  column_title = NULL,
  heatmap_legend_param = list(
    at = c(-2, 0, 2),
    labels = c("-2", "0", "2"),
    title_gp = gpar(fontsize = 12),
    labels_gp = gpar(fontsize = 11)
  )
)

pdf(file.path(fig_dir, "tsoi_marker_genes_complexheatmap.pdf"),
    width = 5.5, height = 10)
draw(ht)
dev.off()

draw(ht)
```


```{r}
TOP_MARKERS_PER_CLUSTER <- 5
RNA_MARKER_LOGFC_CUTOFF <- 0.5
RNA_MARKER_PADJ_CUTOFF <- 0.05

# Filter less-interpretable transcript names.
# This is not perfect, but it helps make the dot plot readable.
unannotated_pattern <- "^(LINC|LOC|MIR|SNOR|RNU)\\d|^A[CFLP]\\d{6}|\\.\\d+$"

top_rna_markers <- rna_all_markers %>%
  filter(
    p_val_adj < RNA_MARKER_PADJ_CUTOFF,
    avg_log2FC > RNA_MARKER_LOGFC_CUTOFF,
    !grepl(unannotated_pattern, gene)
  ) %>%
  mutate(
    specificity = avg_log2FC * (pct.1 - pct.2)
  ) %>%
  group_by(cluster) %>%
  slice_max(specificity, n = TOP_MARKERS_PER_CLUSTER, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(as.numeric(as.character(cluster)), desc(specificity))

write.csv(
  top_rna_markers,
  file.path(tbl_dir, "top_RNA_markers_for_dotplot_specificity.csv"),
  row.names = FALSE
)

top_rna_genes <- unique(top_rna_markers$gene)

cat("Top marker genes selected for dot plot:", length(top_rna_genes), "\n")
print(top_rna_markers)

DefaultAssay(merged) <- "RNA"
Idents(merged) <- "wnn_clusters"

top_rna_genes <- intersect(top_rna_genes, rownames(merged[["RNA"]]))

if (length(top_rna_genes) == 0) {
  stop("None of the selected marker genes were found in the RNA assay.")
}

p_top_rna_markers <- DotPlot(
  merged,
  features = top_rna_genes,
  group.by = "wnn_clusters",
  assay = "RNA"
) +
  scale_color_gradient2(
    low = "#4575b4",
    mid = "white",
    high = "#d73027",
    midpoint = 0
  ) +
  labs(
    x = "Top RNA marker genes",
    y = "WNN cluster",
    color = "Average\nexpression",
    size = "Percent\nexpressed",
    title = "Top RNA markers define WNN cluster identities"
  ) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(
      size = 10,
      angle = 90,
      hjust = 1,
      vjust = 0.5,
      face = "italic"
    ),
    axis.text.y = element_text(size = 11),
    plot.title = element_text(face = "bold")
  )

print(p_top_rna_markers)

ggsave(
  file.path(fig_dir, "top_RNA_markers_dotplot_specificity.pdf"),
  p_top_rna_markers,
  width = max(10, 0.35 * length(top_rna_genes)),
  height = 5.5
)

cat(
  "Top marker dot plot saved to:",
  file.path(fig_dir, "top_RNA_markers_dotplot_specificity.pdf"),
  "\n"
)
```

